---
title: "Predicting Manner of Exercising from Accelerometer Measurements Using Machine  Learning"
author: "Sakthi"
output:
output: pdf_document
---

## Background
This project aims to predict in which manner participants completed a single exercise, based on recordings provided by accelerometers attached to 4 measurement points on the body. Six male participants aged 20-28 were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in one correct, and 4 incorrect fashions. Specifically, the exercises were performed:  
* Exactly according to the specification (__Class A__);  
* Throwing elbows to the front (__Class B__);  
* Lifting the dumbbell only halfway (__Class C__);  
* Lowering the dumbbell only halfway (__Class D__); and  
* Throwing the hips to the front (__Class E__).

Measurements were taken using four 9 degrees of freedom Razor inertial measurement units (IMU), which were attached to the participant's chest, upper arm and forearm (glove) and the dumbbell (Figure 1). To build the study features from these devices, the authors of the study used a sliding window approach with different lengths from 0.5 second to 2.5 seconds, with 0.5 second overlap. In each step of the sliding window approach they calculated features on the Euler angles (roll, pitch and yaw), as well as the raw accelerometer, gyroscope and magnetometer readings. For the Euler angles of each of the four sensors they calculated eight features: mean, variance, standard deviation, max, min, amplitude, kurtosis and skewness.

Set working directory uding setwd command prior to executig the script
## Setup required library
```{r  message = FALSE, cache = TRUE}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
library(corrplot)
library(gbm)
```

## Loading in the data and splitting into training and testing sets


```{r  message = FALSE, cache = TRUE}
setwd("F:/Download")
train_in <- read.csv('pml-training.csv', header=T)
valid_in <- read.csv('pml-testing.csv', header=T)
dim(train_in)

dim(valid_in)
```

## Cleaning the input data
```{r  message = FALSE, cache = TRUE}
trainData<- train_in[, colSums(is.na(train_in)) == 0]
validData <- valid_in[, colSums(is.na(valid_in)) == 0]
dim(trainData)
dim(validData)
```
## We will remove the first seven variables as they have little impact on the outcome 
```{r  message = FALSE, cache = TRUE}
trainData <- trainData[, -c(1:7)]
validData <- validData[, -c(1:7)]
dim(trainData)
dim(validData)
```
###Preparing the datasets for prediction

Preparing the data for prediction by splitting the training data into 70% as train data and 30% as test data. This splitting will server also to compute the out-of-sample errors.

The test data renamed: valid_in (validate data) will stay as is and will be used later to test the prodction algorithm on the 20 cases.
```{r  message = FALSE, cache = TRUE}
set.seed(1234) 
inTrain <- createDataPartition(trainData$classe, p = 0.7, list = FALSE)
trainData <- trainData[inTrain, ]
testData <- trainData[-inTrain, ]
dim(trainData)
dim(testData)
```

## Removing the variables that are near-zero-variance
```{r  message = FALSE, cache = TRUE}
NZV <- nearZeroVar(trainData)
trainData <- trainData[, -NZV]
testData  <- testData[, -NZV]
dim(trainData)
dim(testData)
dim(testData)
```
After this cleaning we are down now to 53 variables

The following correlation plot uses the following parameters (source:CRAN Package ‘corrplot’) “FPC”: the first principal component order. “AOE”: the angular order tl.cex Numeric, for the size of text label (variable names) tl.col The color of text label.
```{r  message = FALSE, cache = TRUE}
cor_mat <- cor(trainData[, -53])
corrplot(cor_mat, order = "FPC", method = "color", type = "upper", 
         tl.cex = 0.8, tl.col = rgb(0, 0, 0))
```
we use the findCorrelation function to search for highly correlated attributes with a cut off equal to 0.75

```{r  message = FALSE, cache = TRUE}
highlyCorrelated = findCorrelation(cor_mat, cutoff=0.75)
```
We then obtain the names of highly correlated attributes
```{r  message = FALSE, cache = TRUE}
names(trainData)[highlyCorrelated]
```

##Model building

For this project we will use two different algorithms, classification trees and random forests, to predict the outcome.

    classification trees
    random forests
    Generalized Boosted Model

##Prediction with classification trees

We first obtail the model, and then we use the fancyRpartPlot() function to plot the classification tree as a dendogram.
```{r  message = FALSE, cache = TRUE}
set.seed(12345)
decisionTreeMod1 <- rpart(classe ~ ., data=trainData, method="class")
fancyRpartPlot(decisionTreeMod1)
```
We then validate the model “decisionTreeModel” on the testData to find out how well it performs by looking at the accuracy variable.
```{r  message = FALSE, cache = TRUE}
predictTreeMod1 <- predict(decisionTreeMod1, testData, type = "class")
cmtree <- confusionMatrix(predictTreeMod1, testData$classe)
cmtree
```
plot matrix results
```{r  message = FALSE, cache = TRUE}
plot(cmtree$table, col = cmtree$byClass, 
     main = paste("Decision Tree - Accuracy =", round(cmtree$overall['Accuracy'], 4)))
```

We see that the accuracy rate of the model is low: 0.6967 and therefore the out-of-sample-error is about 0.3 which is considerable.

##Prediction with Random Forest

We first determine the model
```{r  message = FALSE, cache = TRUE}
controlRF <- trainControl(method="cv", number=3, verboseIter=FALSE)
modRF1 <- train(classe ~ ., data=trainData, method="rf", trControl=controlRF)
modRF1$finalModel
```
We then validate the model obtained model “modRF1” on the test data to find out how well it performs by looking at the Accuracy variable
```{r  message = FALSE, cache = TRUE}
predictRF1 <- predict(modRF1, newdata=testData)
cmrf <- confusionMatrix(predictRF1, testData$classe)
cmrf
```
The accuracy rate using the random forest is very high: Accuracy : 1 and therefore the out-of-sample-error is equal to 0***. But it might be due to overfitting.

Let’s plot the model

```{r  message = FALSE, cache = TRUE}
plot(modRF1)

plot(cmrf$table, col = cmrf$byClass, main = paste("Random Forest Confusion Matrix: Accuracy =", round(cmrf$overall['Accuracy'], 4)))
```
##Prediction with Generalized Boosted Regression Models
```{r  message = FALSE, cache = TRUE}
set.seed(12345)
controlGBM <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
modGBM  <- train(classe ~ ., data=trainData, method = "gbm", trControl = controlGBM, verbose = FALSE)
modGBM$finalModel
```
```{r  message = FALSE, cache = TRUE}
print(modGBM)
```
## Validate the GBM model 
```{r  message = FALSE, cache = TRUE}
predictGBM <- predict(modGBM, newdata=testData)
cmGBM <- confusionMatrix(predictGBM, testData$classe)
cmGBM
```

The accuracy rate using the random forest is very high: Accuracy : 0.9736 and therefore the *out-of-sample-error is equal to 0.0264**.
Applying the best model to the validation data

By comparing the accuracy rate values of the three models, it is clear the the ‘Random Forest’ model is the winner. So will use it on the validation data
```{r  message = FALSE, cache = TRUE}
Results <- predict(modRF1, newdata=validData)
Results
```